<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{urltools}
-->

##Elegant URL handling with urltools
URLs are treated, by base R, as nothing more than components of a data retrieval process: they exist
to create connections to retrieve datasets. This is an essential feature for the language to have,
but it also means that URL handlers are designed for situations where URLs *get* you to the data - 
not situations where URLs *are* the data.

There is no support for encoding or decoding URLs en-masse, and no support for parsing and
interpreting them. `urltools` provides this support!

### URL encoding and decoding

Base R provides two functions - <code>URLdecode</code> and <code>URLencode</code> - for taking percentage-encoded
URLs and turning them into regular strings, or vice versa. As discussed, these are primarily designed to
enable connections, and so they have several inherent limitations, including a lack of vectorisation, that
make them unsuitable for large datasets.

Not only are they not vectorised, they also have several particularly idiosyncratic bugs and limitations:
<code>URLdecode</code>, for example, breaks if the decoded value is out of range:

```{r, eval=FALSE}
URLdecode("test%gIL")
Error in rawToChar(out) : embedded nul in string: '\0L'
In addition: Warning message:
In URLdecode("%gIL") : out-of-range values treated as 0 in coercion to raw
```

URLencode, on the other hand, encodes slashes on its most strict setting - without
paying attention to where those slashes *are*: if we attempt to URLencode an entire URL, we get:

```{r, eval=FALSE}
URLencode("https://en.wikipedia.org/wiki/Article", reserved = TRUE)
[1] "https%3a%2f%2fen.wikipedia.org%2fwiki%2fArticle"
```
That's a completely unusable URL (or ewRL, if you will).

urltools replaces both functions with <code>url\_decode</code> and <code>url\_encode</code> respectively:
```{r, eval=FALSE}
library(urltools)
url_decode("test%gIL")
[1] "test"
url_encode("https://en.wikipedia.org/wiki/Article")
[1] "https://en.wikipedia.org%2fwiki%2fArticle"
```

As you can see, <code>url\_decode</code> simply excludes out-of-range characters from consideration, while <code>url\_encode</code> detects characters that make up part of the URLs scheme, and leaves them unencoded. Both are extremely fast; with `urltools`, you can
decode a vector of 1,000,000 URLs in 0.9 seconds.

### URL parsing

Once you've got your nicely decoded (or encoded) URLs, it's time to do something with them - and, most of the time,
you won't actually care about most of the URL. You'll want to look at the scheme, or the domain, or the path,
but not the entire thing as one string.

The solution is <code>url_parse</code>, which takes a URL and breaks it out into its [RfC 3986](http://www.ietf.org/rfc/rfc3986.txt) components: scheme, domain, port, path, query string and fragment identifier. This is,
again, fully vectorised, and can happily be run over hundreds of thousands of URLs, rapidly processing them. The
results are provided as a data.frame, since most people use data.frames to store data.

```{r, eval=FALSE}
> parsed_address <- url_parse("https://en.wikipedia.org/wiki/Article")
> str(parsed_address)
'data.frame':	1 obs. of  6 variables:
 $ scheme   : chr "https"
 $ domain   : chr "en.wikipedia.org"
 $ port     : chr ""
 $ path     : chr "wiki/article"
 $ parameter: chr ""
 $ fragment : chr ""                           
```

We can also perform the opposite of this operation with `url_compose`:
```{r, eval=FALSE}
> url_compose(parsed_address)
[1] "https://en.wikipedia.org/wiki/article"
```

### Value extraction
Once a URL is parsed, it's sometimes useful to get the value associated with a particular query parameter. As
an example, take the URL `http://en.wikipedia.org/wiki/api.php?action=parse&pageid=1023&export=json`. What
pageID is being used? What is the export format? We can find out with `param_get`.

```{r, eval=FALSE}
> str(param_get(urls = "http://en.wikipedia.org/wiki/api.php?action=parse&pageid=1023&export=json",
                     parameter_names = c("pageid","export")))
'data.frame':	1 obs. of  2 variables:
 $ pageid: chr "1023"
 $ export: chr "json"
```

### Getting/setting URL components
With the inclusion of a URL parser, we suddenly have the opportunity for lubridate-style component getting
and setting. Syntax is identical to that of `lubridate`, but uses URL components as function names.

```{r, eval=FALSE}
url <- "https://en.wikipedia.org/wiki/Article"
scheme(url)
"https"
scheme(url) <- "ftp"
url
"ftp://en.wikipedia.org/wiki/Article"
```
Fields that can be extracted or set are <code>scheme</code>, <code>domain</code>, <code>port</code>, <code>path</code>,
<code>parameters</code> and <code>fragment</code>.

### Suffix extraction

Once we've extracted a domain from a URL with `domain` or `url_parse`, we can identify which bit is the domain name, and which
bit is the "top-level domain" or suffix, thanks to code contributed by Jay Jacobs:

```{r, eval=FALSE}
> url <- "https://en.wikipedia.org/wiki/Article"
> domain_name <- domain(url)
> domain_name
[1] "en.wikipedia.org"
> str(suffix_extract(domain_name))
'data.frame':	1 obs. of  4 variables:
 $ host     : chr "en.wikipedia.org"
 $ subdomain: chr "en"
 $ domain   : chr "wikipedia"
 $ tld      : chr "org"
```

### Other URL handlers
If you have ideas for other URL handlers that would make your data processing easier, the best approach
is to either [request it](https://github.com/Ironholds/urltools/issues) or [add it](https://github.com/Ironholds/urltools/pulls)!
